A study on the usage of the wireless signal spectrum in a City of Things network

### 2017-05-15 advnetwpaper/paper_no_ieeetrans.tex
Added script to plot multiple days' smoothings near to eachother (did not replot raw as that would serve no purpose and be hard to interpret). Made version of paper without ieeetrans. See file plot_week_smoothings.py for a command to run it.

%% bare_conf.tex
%% V1.4b
%% 2015/08/26
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
%% conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/pkg/ieeetran
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE!
%% User assumes all risk.
%% In no event shall the IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

\documentclass[conference]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

@@

% *** CITATION PACKAGES ***
%
\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.

\usepackage{siunitx}
\usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
\graphicspath{{../images/}}

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at:
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Bare Demo of IEEEtran.cls\\ for IEEE Conferences}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Michael Shell}
\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
Georgia Institute of Technology\\
Atlanta, Georgia 30332--0250\\
Email: http://www.michaelshell.org/contact.html}
\and
\IEEEauthorblockN{Homer Simpson}
\IEEEauthorblockA{Twentieth Century Fox\\
Springfield, USA\\
Email: homer@thesimpsons.com}
\and
\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
\IEEEauthorblockA{Starfleet Academy\\
San Francisco, California 96678--2391\\
Telephone: (800) 555--1212\\
Fax: (888) 555--1212}}

@@

% no keywords

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
% no \IEEEPARstart
Most of the radio spectrum requires a license to operate on. Devices emitting Radio Frequency (RF) energy without a license are limited to using one of the unlicensed radio bands, known as the Industrial, Scientific and Medical (ISM) radio bands. The most well-known of these bands is the 100MHz wide band centered around 2.45GHz, often called the 2.4GHz band. Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4GHz band. Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4GHz band, along with the 5GHz band (which actually ranges from 5.725GHz to 5.875GHz). These ISM bands were originally intended and still used for devices emitting RF energy for purposes other than communication. The most common such application is the microwave oven. All this leads to high interference in these bands. In this paper, we investigate how bad this 2.4GHz and 5GHz interference actually is in the city of Antwerp. \\ \\
To measure this interference we use the City of Things (CoT) network, operated by research institute imec. This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp. On TODO (20?) of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 10 minutes for TODO days. The main motivation of this paper was in fact to investigate whether interference around the gateways is significant enough to pose a problem for the network. Next to the general magnitude of the interference, we look for patterns in daily and weekly variation of the interference.

@@

\hfill mds

\hfill August 26, 2015

@@

The per-bin magnitudes indicate how power is divided within the channel. To understand these values, some background about radio signaling is needed. Signals in radio communication are usually created using a technique called \textit{IQ modulation}. With this technique, the final signal is a combination of two separate signals (the in-phase or \textit{I} and the quadrature or \textit{Q} signals) whose phases are exactly 90 degrees apart. By changing only the amplitude of the \textit{I} and \textit{Q} signals, the resulting signal's amplitude and phase can attain any value. The reported bin magnitude $b(i)$ for bin $i$ is the sum of the absolute values of the \textit{I} and \textit{Q} signals' magnitudes. This magnitude in turn is the amplitude squared. The per-bin power scales quadratically with this magnitude value: if bin $i$ has twice the magnitude of bin $j$, bin $i$'s share of the channel's power is four times as large as bin $j$'s. We can compute a coefficient $c(i)$ for every bin $i \in [1,n]$ as follows: $c(i) = \frac{b(i)^2}{\sum\limits_{j=1}^{n}b(j)^2}$. Note that every $c(i) \in [0,1]$ and $\sum\limits_{c=1}^{n}{c(i)} = 1$. \\

@@

>>>>

\documentclass[a4paper, 11pt]{article}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

@@

% *** CITATION PACKAGES ***
%
\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.
\usepackage{siunitx}
\usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
\graphicspath{{../images/}}

% *** MATH PACKAGES ***
%
\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{A study on the usage of the wireless signal spectrum in a City of Things network}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{
	Jakob Struye (s.......)\\ 
    Thomas Hendriks (s0139858)\\
}

@@

\section{Introduction}
% no \IEEEPARstart
Most of the radio spectrum requires a license to operate on. Devices emitting Radio Frequency (RF) energy without a license are limited to using one of the unlicensed radio bands, known as the Industrial, Scientific and Medical (ISM) radio bands. The most well-known of these bands is the 100MHz wide band centered around 2.45GHz, often called the 2.4GHz band. Many consumer devices such as wireless mice and keyboards, bluetooth headphones and Radio Controlled (RC) cars communicate in the 2.4GHz band. Furthermore most IEEE 802.11 (Wi-Fi) routers operate in the 2.4GHz band, along with the 5GHz band (which actually ranges from 5.725GHz to 5.875GHz). These ISM bands were originally intended and still used for devices emitting RF energy for purposes other than communication. The most common such application is the microwave oven. All this leads to high interference in these bands. In this paper, we investigate how bad this 2.4GHz and 5GHz interference actually is in the city of Antwerp. \\ \\
To measure this interference we use the City of Things (CoT) network, operated by research institute imec. This network consists of hundreds of sensors and wireless gateways positioned around the city of Antwerp. On 13 of these gateways, we sampled activity on the 2.4GHz and 5GHz bands every 10 minutes for TODO days. The main motivation of this paper was in fact to investigate whether interference around the gateways is significant enough to pose a problem for the network. Next to the general magnitude of the interference, we look for patterns in daily and weekly variation of the interference.

@@

@@

The per-bin magnitudes indicate how power is divided within the channel. To understand these values, some background about radio signaling is needed. Signals in radio communication are usually created using a technique called \textit{IQ modulation}. With this technique, the final signal is a combination of two separate signals (the in-phase or \textit{I} and the quadrature or \textit{Q} signals) whose phases are exactly 90 degrees apart. By changing only the amplitude of the \textit{I} and \textit{Q} signals, the resulting signal's amplitude and phase can attain any value. The reported bin magnitude $b(i)$ for bin $i$ is the sum of the absolute values of the \textit{I} and \textit{Q} signals' magnitudes. This magnitude in turn is the amplitude squared. The per-bin power scales quadratically with this magnitude value: if bin $i$ has twice the magnitude of bin $j$, bin $i$'s share of the channel's power is four times as large as bin $j$'s. We can compute a coefficient $c(i)$ for every bin $i \in [1,n]$ as follows:\\\\\hspace*{1.5em}$c(i) = \dfrac{b(i)^2}{\sum\limits_{j=1}^{n}b(j)^2}$ (Note that every $c(i) \in [0,1]$ and $\sum\limits_{c=1}^{n}{c(i)} = 1$.) \\\\

@@

\section{Measurements}
At first we measured the spectrum on every channel in both the 2.4GHz and the 5GHz band every 10 minutes. Looking at interpretations of the resulting data it was clear that a shorter time between scans would be useful because recurring spikes were missed on some days while they definitely there and also quite interesting.
\\< images waarop da duidelijk is >\\
We then reduced the time between scans from 10 minutes to 2 minutes, which was the minimum we could achieve while guaranteeing that scans would be succesful.
Using an adapted version of Simon Wünderlich's fft_eval code, GNUplot and Savitzky-Golay (source? TODO) filtering we graphically represent our data points per node.
<tussen maandag 15 tem zondag 21 mei>

\section{Findings}
